---
title: "Introduction to the Starlng package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Starlng package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 


```{r libraries, message=FALSE, warning=FALSE}
library(ClustAssess)
library(Starlng)
library(monocle3)
library(Seurat)
library(SeuratData)
library(ggplot2)
library(dplyr)
library(patchwork)
```

# Presenting the Data
In this vignette we will demonstrate how to apply the main Starlng functionalities to single-cell RNA-seq data. We will exemplify this on the PBMC3k dataset.
```{r install_pbmc3k, message=FALSE, warning=FALSE}
InstallData("pbmc3k")
data("pbmc3k")

n_cores <- 4
```

We repeated the procedures of downloading, qc filtering, normalising and dimensionality reduction as described in the `ClustAssess` [vignette](https://github.com/Core-Bioinformatics/ClustAssess/blob/main/vignettes/stability-pipeline-description.Rmd).
```{r basic_seurat_processing, message=FALSE, warning=FALSE}
pbmc3k <- UpdateSeuratObject(pbmc3k)
pbmc3k <- PercentageFeatureSet(pbmc3k, pattern = "^MT-", col.name = "percent.mito")
pbmc3k <- PercentageFeatureSet(pbmc3k, pattern = "^RP[SL][[:digit:]]", col.name = "percent.rp")
all.index <- seq_len(nrow(pbmc3k))
MT.index <- grep(pattern = "^MT-", x = rownames(pbmc3k), value = FALSE)
RP.index <- grep(pattern = "^RP[SL][[:digit:]]", x = rownames(pbmc3k), value = FALSE)
pbmc3k <- pbmc3k[!((all.index %in% MT.index) | (all.index %in% RP.index)), ]
pbmc3k <- subset(
    pbmc3k,
    nFeature_RNA < 2000 & nCount_RNA < 2500 &
    percent.mito < 7 & percent.rp > 7
)
pbmc3k <- NormalizeData(pbmc3k, verbose = FALSE)
pbmc3k <- FindVariableFeatures(
    pbmc3k,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
)
pbmc3k <- ScaleData(pbmc3k, features = rownames(pbmc3k), verbose = FALSE)
pbmc3k <- RunPCA(pbmc3k,
    npcs = 30,
    approx = FALSE,
    verbose = FALSE
)
set.seed(42)
pbmc3k <- RunUMAP(pbmc3k,
    reduction = "pca",
    dims = 1:30,
    n.neighbors = 30,
    min.dist = 0.3,
    metric = "cosine",
    verbose = FALSE
)
pbmc3k <- FindNeighbors(pbmc3k, reduction = "pca", dims = 1:30, verbose = FALSE)
pbmc3k <- FindClusters(pbmc3k, resolution = 1, verbose = FALSE)
pbmc3k
```

The cells are already annotated, as shown in the following UMAP. The annotations can be used to confirm the downstream analysis.
```{r umap_annotations, fig.width=7, fig.height=7}
DimPlot(pbmc3k, reduction = "umap", group.by = "seurat_annotations")
```

# Running Starlng
## Trajectory Inference and Autocorrelation Scoring

Starlng relies on the main principles from the Monocle3 package. Therefore, the
first steps consist in converting our object to Monocle3 and apply the first
processing steps. For conversion, we have used the `create_monocle_default`
function from the `ClustAssess` package.
```{r create_monocle_obj, fig.width=7, fig.height=7}
monocle_obj <- create_monocle_default(
    normalized_expression_matrix = GetAssayData(pbmc3k, assay = "RNA", layer = "data"),
    metadata_df = pbmc3k@meta.data,
    pca_embedding = Embeddings(pbmc3k, reduction = "pca"),
    umap_embedding = Embeddings(pbmc3k, reduction = "umap")
)
plot_cells(
    monocle_obj,
    color_cells_by = "seurat_annotations",
    cell_size = 1,
    group_label_size = 6
)
```

Once the Monocle3 object is created, we apply the SimplePPT algorithm to infer
the trajectory graph. The method relies on defining a number of center points
that represent each subpopulation. Monocle3 fixes an adaptive policy defined
as $15 \cdot \log_10(ncells)$.
Changing the factor can improve the granularity of the branching, as it can be
seen in the following comparison. Therefore, Starlng provides an wrapper that
allows changing this variable.
```{r impact_nodes_per_log10, fig.width = 8, fig.height = 6}
n_nodes <- c(10, 30)
plot_list <- list()
for (n_node in n_nodes) {
    monocle_obj <- custom_learn_graph(
        mon_obj = monocle_obj,
        nodes_per_log10_cells = n_node,
        learn_graph_controls = list(
            eps = 1e-5,
            maxiter = 100,
            prune_graph = TRUE
        ),
        use_partition = FALSE,
        use_closed_loops = FALSE
    )
    plot_list[[as.character(n_node)]] <- plot_cells(
        monocle_obj,
        color_cells_by = "seurat_annotations",
        cell_size = 1) +
      ggtitle(paste0("Trajectory inferred on\n", n_node, " per log10(#cells)"))
}
wrap_plots(plot_list, ncol = 2)
```
## Pseudotime Ordering

The graph structure allows defining an ordering of the centroid points (and
subsequently of each cell). Given that the graph is undirected, choosing the
starting point can be arbitrary. Monocle3's approach is to let the user select
the root interactively using a ShinyApp.
Starlng introduces an objective approach that determines the root point that
maximises the pseudotime range. We justify this objective understanding that
every minor change between neighbouring subtypes can provide useful insights.
For this, we determine the diameter of the graph, which will represent the main
trajectory of the data. We will select the trajectory leaf that belongs to
the half with a weaker branching effect.
```{r pseudotime_recommendation}
trajectory_object <- get_trajectory_object(monocle_obj)
recommended_pseudotime <- get_pseudotime_recommendation(monocle_obj)
```

Starlng proposes a data-driven way to determine the root point as well. The
user can utilise either a metadata category or a group of genes (using a voting
scheme) and apply Starlng to determine a small set of cells that would represent
the provided population.
By default, the identification of the optimal pseudotime range is accompanied
by a recommendation of the metadata group that can be used to obtain a similar
result. In this case, the CD14+ Mono cells were selected as the most appropriate
starting point.
```{r pseudotime_by_metadata, fig.width = 6, fig.height = 6}
umap_df <- as.data.frame(reducedDims(monocle_obj)$UMAP)
mtd_df <- colData(monocle_obj)
umap_df$pseudotime <- recommended_pseudotime$recommended_pseudotime

recommended_mtd_name <- recommended_pseudotime$recommended_mtd_name
recommended_mtd_group <- recommended_pseudotime$recommended_mtd_group
mtd_group <- rownames(mtd_df)[mtd_df[[recommended_mtd_name]] == recommended_mtd_group]
mtd_group <- mtd_group[!is.na(mtd_group)]
recommended_starting_point <- filter_central_cells_from_group(
    cell_group = mtd_group,
    umap_embedding = umap_df[, c("UMAP_1", "UMAP_2")],
    n_points = 5
)
str(recommended_pseudotime)
ggplot() +
    geom_point(
        data = umap_df,
        aes(x = UMAP_1, y = UMAP_2, color = pseudotime),
        size = 1, alpha = 0.5) +
    theme_classic() +
    scale_color_viridis_c() +
    geom_point(
        data = umap_df[recommended_starting_point, ],
        aes(x = UMAP_1, y = UMAP_2),
        color = "red", size = 3) +
    ggtitle("UMAP Colored by Recommended Pseudotime\n
    (Red cells recommended as Starting Point)") +
    plot_trajectory_graph(trajectory_object, plot_nodes = 0, edge_size = 1)$layers
```

## Autocorrelation Scoring

Another use of the trajectory graph is to describe a space that can be further
utilised to infer autocorrelation patterns of each gene using the Moran's I test.
```{r graph_test, include=FALSE}
invisible(capture.output({
    autocorrelation_result <- monocle3::graph_test(
        cds = monocle_obj,
        neighbor_graph = "principal_graph",
        cores = n_cores,
        verbose = FALSE
    )
}))
```

```{r graph_test_show, eval=FALSE}
autocorrelation_result <- monocle3::graph_test(
    cds = monocle_obj,
    neighbor_graph = "principal_graph",
    cores = n_cores,
    verbose = FALSE
)
```

The Moran's I values range between -1 and 1. Values close to 0 reflect a random
localisation pattern of the cells expressing a gene. Positive values denote that
cells are clustered in more compact groups.
The presence of noise, as well as the plasticity of the cells increases the
difficulty of identifying genes that are well clustered. As it can be observed
in the following plot, the number of genes decreases exponentially as the Moran's
I increases.
```{r moran_i_distribution, fig.width = 6, fig.height=5}
rownames(autocorrelation_result) <- autocorrelation_result$gene_short_name
autocorrelation_result <- autocorrelation_result[!is.na(autocorrelation_result$morans_I), ]
autocorrelation_result <- autocorrelation_result %>%
    dplyr::select(c("gene_short_name", "morans_test_statistic", "morans_I", "q_value")) %>%
    dplyr::arrange(dplyr::desc(.data$morans_I))


morans_I_bins <- seq(-1, 1, by = 0.05)
autocorrelation_result$morans_I_bin <- cut(
    autocorrelation_result$morans_I,
    breaks = morans_I_bins,
    include.lowest = TRUE
)
autocorrelation_result$morans_I_bin <- as.character(autocorrelation_result$morans_I_bin)
autocorrelation_counts <- autocorrelation_result %>%
    group_by(morans_I_bin) %>%
    summarise(count = n())
monocle_obj <- monocle_obj[rownames(autocorrelation_result), ]

ggplot(autocorrelation_counts, aes(x = morans_I_bin, y = count)) +
    geom_histogram(stat = "identity") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_log10() +
    ylab("Number of Genes (log10)") +
    ggtitle("Distribution of Moran's I Values")

head(autocorrelation_result)
```

We define other usual gene statistics, such as average expression, percentage of
cells expressed. The following scatterplot suggests that these metrics can't
act as strong predictors of the autocorrelation score.
```{r add_gene_stats, fig.width = 7, fig.height = 6}
expr_matrix <- monocle_obj@assays@data$normalized_data
autocorrelation_result$average_expression <- Matrix::rowMeans(expr_matrix)
avg_non_zero <- apply(expr_matrix, 1, function(x) mean(x > 0))
autocorrelation_result$average_expression_non_zero <- avg_non_zero
autocorrelation_result$n_expressed_cells <- rowSums(expr_matrix > 0)
perc_cells <- autocorrelation_result$n_expressed_cells / ncol(expr_matrix) * 100
autocorrelation_result$percent_expressed_cells <- perc_cells
autocorrelation_result$p_val <- -log10(autocorrelation_result$q_value + 1e-10)
autocorrelation_result[autocorrelation_result$q_value >= 0.05, "p_val"] <- NA

head(autocorrelation_result)

ggplot(autocorrelation_result, aes(
    x = average_expression_non_zero,
    y = morans_I,
    size = percent_expressed_cells,
    colour = p_val)) +
    geom_point() +
    theme_classic() +
    xlab("Average Non-Zero Expression") +
    ylab("Moran's I") +
    ggtitle("Moran's I vs Average Non-Zero Expression") +
    scale_colour_viridis_c()
```

## Identification of Gene Modules

For the following analysis, we will filter the genes based on the significance
of their autocorrelation patterns. We selected a significance pvalue threshold
of 0.05 and a Moran's I threshold of 0.1, in an attempt to describe a large
enough landscape of signal genes. We note that other threshold I values can be
appropriate and could be investigated in the analysis.

We process the filtered genes by calculating their feature loading after
performing the Principal Component Analysis at cell level.
```{r gene_filtering_processing}
remaining_genes <- autocorrelation_result %>%
    filter(q_value < 0.05 & morans_I > 0.1, ) %>%
    rownames()
gene_loadings <- get_feature_loading(
    expr_matrix = expr_matrix[remaining_genes, ],
    npcs = 30, approx = FALSE)
```

The reduced space is then given as input for the clustering pipeline, which
follows the PhenoGraph framework (also used by default in Monocle3). Additionally,
we run the pipeline by varying the number of neighbours (used to infer the gene
adjacency graph based on neighbourhood overlap) and the resolution parameter
(which directly impacts the number of resulting clusters). Each configuration
is run 100 times with different random seed initialisations.

The pipeline can be run in parallel across all platforms, using PSOCK clusters.
```{r gene_clustering, message=FALSE, warning=FALSE}
if (n_cores > 1) {
    cl <- parallel::makePSOCKcluster(n_cores)
    doParallel::registerDoParallel(cl)
}

clustering_assessment <- clustering_pipeline(
    embedding = gene_loadings,
    n_neighbours = seq(5, 50, by = 5),
    graph_type = "snn",
    prune_value = -1,
    resolutions = list(
        "RBConfigurationVertexPartition" = seq(from = 0.1, to = 2, by = 0.1)
    ),
    number_iterations = 5,
    number_repetitions = 100,
    merge_identical_partitions = TRUE
)

if (n_cores > 1) {
    parallel::stopCluster(cl)
    foreach::registerDoSEQ()
}
```

We utilise the clustering assessment principles from ClustAssess to infer the
robustness of results for the different number of nearest neighbours. We use the
same two-step approach, that firstly keeps the upper half of configurations ordered
by the median Element-Centric Consistency (ECC) and the selects the configuration
with the lowest ECC IQR.
```{r best_nn_clustering, fig.width = 7, fig.height = 5}
per_n_neigh_ecc <- do.call(rbind,
    lapply(names(clustering_assessment$clusters_list), function(n_neigh) {
        x <- clustering_assessment$clusters_list[[n_neigh]]
        data.frame(
            ECC = x$RBConfigurationVertexPartition$overall_ecc,
            n_neighbours = n_neigh
        )
}))
per_n_neigh_ecc$n_neighbours <- factor(
    per_n_neigh_ecc$n_neighbours,
    levels = as.character(seq(5, 50, by = 5))
)
per_n_neigh_stats <- per_n_neigh_ecc %>%
    group_by(n_neighbours) %>%
    summarise(
        median_ecc = median(ECC),
        iqr_ecc = IQR(ECC)
    )
# lowest median from the first half
median_threshold <- sort(per_n_neigh_stats$median_ecc)[ceiling(nrow(per_n_neigh_stats) / 2)]
iqr_text <- per_n_neigh_stats %>%
    filter(median_ecc >= median_threshold) %>%
    arrange(iqr_ecc) %>%
    mutate(
        iqr_text = paste0("IQR: ", round(iqr_ecc, 3)),
        text_colour = ifelse(iqr_ecc == min(iqr_ecc), "#519451", "black")
    )
max_y <- max(per_n_neigh_ecc$ECC) * 1.01

ggplot(per_n_neigh_ecc, aes(x = n_neighbours, y = ECC)) +
    geom_boxplot() +
    theme_classic() +
    xlab("Number of Neighbours") +
    ylab("Overall ECC") +
    geom_hline(yintercept = median_threshold, linetype = "dashed", color = "red") +
    geom_text(
        data = iqr_text,
        aes(x = n_neighbours, y = max_y, label = iqr_text, color = text_colour),
        vjust = -0.5,
        size = 3
    ) +
    scale_color_identity() +
    ggtitle("ECC vs Number of Neighbours")

best_configuration <- select_best_configuration(clustering_assessment$clusters_list)
print(best_configuration)
```

For this data, the optimal number of nearest neighbours is 20. We fix these
parameters and calculate the gene UMAP.
```{r update_gene_umap_and_clusters}
best_nn <- best_configuration[[1]]
best_quality <- best_configuration[[2]]
gene_adj_matrix <- clustering_assessment$embedding_list$adj_matrix[[best_nn]]
set.seed(42)
gene_umap <- uwot::umap(
    gene_loadings,
    n_neighbors = as.numeric(best_nn),
    metric = "cosine",
    n_components = 2,
    verbose = FALSE
)
gene_umap <- as.data.frame(gene_umap)
colnames(gene_umap) <- paste0("Gene_UMAP_", 1:2)

clusters_list <- clustering_assessment$clusters_list[[best_nn]][[best_quality]]
```

The last step of the clustering assessment is to infer the number of modules
that produce similar results across multiple runs. For this, we consider a
number of gene clusters (defined as modules) to be stable if the average ECC
score is greater than 0.9 (from a maximum of 1) and if its associated partition
appear at least 30 times (to ensure a statistically-relevant sample of observations).
```{r best_n_modules, fig.width = 8, fig.height = 5}
per_k_ecc <- do.call(rbind, lapply(names(clusters_list$k), function(k) {
    x <- clusters_list$k[[k]]
    data.frame(
        ECC = x$ecc,
        k = k
    )
}))
per_k_ecc$k <- factor(per_k_ecc$k, levels = names(clusters_list$k))
average_ecc_per_k <- per_k_ecc %>%
    group_by(k) %>%
    summarise(average_ecc = mean(ECC))
max_y <- max(per_k_ecc$ECC) * 1.03
average_ecc_per_k$frequency <- sapply(names(clusters_list$k), function(k) { 
    sum(sapply(clusters_list$k[[k]]$partitions, function(partition) partition$freq))
})
average_ecc_per_k_text <- average_ecc_per_k %>%
    arrange(frequency) %>%
    filter(average_ecc >= 0.9) %>%
    mutate(
        frequency_text = paste0("Freq:\n", frequency),
        text_colour = ifelse(frequency < 30, "black",  "#519451")
    )

ggplot(per_k_ecc, aes(x = k, y = ECC)) +
    geom_boxplot() +
    geom_point(
        data = average_ecc_per_k,
        aes(x = k, y = average_ecc),
        color = "blue",
        size = 2, shape = 22) +
    theme_classic() +
    xlab("k") +
    ylab("ECC") +
    geom_hline(yintercept = 0.9, linetype = "dashed", color = "red") +
    geom_text(
        data = average_ecc_per_k_text,
        aes(x = k, y = max_y, label = frequency_text, color = text_colour),
        vjust = -0.5,
        size = 3
    ) +
    scale_color_identity() +
    coord_cartesian(ylim = c(0.15, 1.09))
    ggtitle("ECC vs k for the Best Configuration")

stable_clusters <- ClustAssess::choose_stable_clusters(
    clusters_list = clusters_list$k,
    ecc_threshold = 0.9,
    freq_threshold = 30,
    summary_function = mean
)

print(names(stable_clusters))
```

We notice that the highest number of modules is 14. We will select this value
for the following analysis.
```{r module_selection}
chosen_n_modules <- names(stable_clusters)[length(stable_clusters)]
module_names <- as.character(seq_len(chosen_n_modules))
chosen_clustering <- setNames(
    stable_clusters[[chosen_n_modules]]$partitions[[1]]$mb,
    remaining_genes
)
module_colours <- setNames(qualpalr::qualpal(length(module_names))$hex, module_names)
module_list <- split(names(chosen_clustering), chosen_clustering)
print(summary(factor(chosen_clustering)))
```

# Module Analysis
## Voting Scheme

We can determine the population described by each gene module using a voting
scheme approach. This voting scheme selects the cells that express at least
`n_coexpressed_thresh` genes above a given threshold (alternative, a percentile
can be provided to infer a dynamic threshold).
In this first example, we select all cells that express at least one gene from
the module. We remark that the voting scheme has been too lenient.
```{r module_binary_selection, fig.width = 8, fig.height = 16}
module_binary_voting <- do.call(cbind, lapply(module_names, function(module) {
    genes_in_module <- names(chosen_clustering)[chosen_clustering == module]
    if (length(genes_in_module) == 0) {
        return(rep(NA, ncol(expr_matrix)))
    }
    as.character(voting_scheme(
        expression_matrix = expr_matrix[genes_in_module, , drop = FALSE],
        thresh_percentile = 0,
        thresh_value = 0,
        n_coexpressed_thresh = 1,
        summary_function = NULL 
    ))
}))
colnames(module_binary_voting) <- module_names

plot_umap_gene_modules(
    module_summaries = module_binary_voting,
    umap_embedding = umap_df[, c("UMAP_1", "UMAP_2")],
    trajectory_object = trajectory_object,
    n_columns = 3,
    combine_legend = TRUE,
    scale_values = FALSE,
    continuous_colors = cont_colours,
    show_labels = FALSE,
    legend_text_size = 20,
    legend_key_size = 5
)
```

The second approach would be to aggregate the expression across all genes belonging
to one module. For comparability across module, we scale the values between 0
and 1. This approach reveal clearer patterns for each module.
```{r module_scaled_average, fig.width=8, fig.height = 16}
module_avg_expression <- do.call(cbind, lapply(module_names, function(module) {
    genes_in_module <- names(chosen_clustering)[chosen_clustering == module]
    if (length(genes_in_module) == 0) {
        return(rep(NA, ncol(expr_matrix)))
    }
    expr_value <- voting_scheme(
        expression_matrix = expr_matrix[genes_in_module, , drop = FALSE],
        thresh_percentile = 0,
        thresh_value = 0,
        n_coexpressed_thresh = 1,
        summary_function = mean
    )
    (expr_value - min(expr_value)) / (max(expr_value) - min(expr_value))
}))
colnames(module_avg_expression) <- module_names

white_red_colours <- c("grey85", "#FFF7EC", "#FEE8C8", "#FDD49E",
"#FDBB84", "#FC8D59", "#EF6548", "#D7301F", "#B30000", "#7F0000")
plot_umap_gene_modules(
    module_summaries = module_avg_expression,
    umap_embedding = umap_df[, c("UMAP_1", "UMAP_2")],
    trajectory_object = trajectory_object,
    n_columns = 3,
    combine_legend = TRUE,
    scale_values = TRUE,
    continuous_colors = white_red_colours,
    show_labels = FALSE,
    legend_text_size = 20,
    colourbar_height = 200,
    cell_size = 1,
    cell_sort_order = "highest"
)
```

Starlng provides a procedure that identifies the core cells that define the
module-specific population. For this, we can either use a threshold for the
scaled values (defaults to 0.5) or a percentage of the top cells, ordered by
the average expression.
```{r module_mask}
scale_threshold <- 0.5
psd_mask <- !is.na(umap_df$pseudotime)
module_avg_expression <- split(module_avg_expression, col(module_avg_expression))
names(module_avg_expression) <- module_names
module_mask <- build_module_masks(
    module_summ = module_avg_expression,
    psd_mask = psd_mask,
    scale_threshold = scale_threshold,
    top_cells_percent = 100
)
names(module_mask) <- module_names
```

## Generating Module Statistics

The previously-determined populations can be further used to generated per-module
stats, such as the position across the pseudotime trajectory, or the average
UMAP intra-cluster distance.
```{r module_stats}
modules_stats <- get_module_stats(
    module_summ = module_avg_expression,
    module_mask = module_mask,
    psd_value = umap_df$pseudotime,
    umap_df = umap_df[, c("UMAP_1", "UMAP_2")],
    centroid = TRUE
)
modules_stats_summary <- summarise_module_stats(
    modules_stats = modules_stats,
    gene_modules = module_list
)
head(modules_stats_summary)
```

Mapping against present metadata
We can also visualise the overlap between the module-induced populations and
existing metadata. In the following bubbleplot we notice that some modules
are specific to a certain group (for example, module 1 to B cells or module 4
to NK cells), while some modules are rather scattered across multiple clusters
(for example, module 8 is highly expressed in B cells, Naive CD4 T cells, Memory
CD4 T cells).
```{r mapping_metadata_bubbleplot, fig.width = 7, fig.height = 5}
mtd_value <- as.character(monocle_obj@colData[, "seurat_annotations"])
mtd_order <- order_metadata_groups_by_pseudotime(mtd_value, umap_df$pseudotime)
plot_module_pseudobulk_expression(
  module_expression_summary = module_avg_expression,
  cell_metadata = mtd_value,
  mtd_order = rev(mtd_order),
  scale = TRUE,
  cap_value = 2.5
)
```

## Detect Outlier Modules

The previous remark implies that some modules might be too scattered for them
to provide relevant biological insights.
Based on this observation, we devised a method that aims to detect outlier
modules by analysing the pseudotime IQR and average UMAP distance patterns.
Modules that have an absolute MAD Z score for both these metrics will be flagged
as outliers.
Supplimentary, we try to compare the performance of the modules against a baseline.
For this, we calculate the value of these two metrics by considering the entire
dataset being a single population. If a module lies in 85% of the baseline values,
we conclude that it's behaviour isn't too different than a random assignation,
hence the module is considered an outlier.

Furthermore, we introduce an intermediary category of _redundant_ modules. Some
gene clusters that were not flagged as outlier are still spanning across a
high range of the pseudotime landscape. Usually, the area covered by these clusters
are better described by multiple modules that represent smaller portions.
To determine the redundancy, we sort the modules based on their pseudotime IQR
(or the average UMAP distance). If a module cover a new part of the dataset with
at least half of its associated cells, it will be considered non-redundant.
Otherwise, we calculate a harmonic mean of the added coverage wrt the entire
dataset and the proportion of cells from the module-specific population that are
involved in this process. If this score is worse than the median of the scores
of the modules already defined as non-redundant, the module will be classified
as redundant.
In our case, the following plot shows that module 2 was added despite its high
IQR because it covers a significant amount of population that has not been
described by the previous modules.
```{r outlier_detection, fig.width = 8, fig.height = 6}
umap_median_dist <- median(calculate_umap_average_distance(umap_df[, 1:2]))
outlier_assessment <- detect_outlier(
    modules_stats = modules_stats_summary,
    cell_masks = module_mask,
    psd_value = umap_df$pseudotime,
    umap_dist_threshold = umap_median_dist * 0.85
)

coverage_df <- outlier_assessment$coverage_evolution_df %>% arrange(module_iqr)
coverage_df$cumulative_coverage <- cumsum(coverage_df$coverage_added)
ggplot(coverage_df, aes(
    x = module_iqr,
    y = cumulative_coverage,
    size = f1_score, colour = median_umap_distance)) +
    geom_point() +
    geom_line(colour = "black", linewidth = 1) +
    theme_classic() +
    xlab("Module IQR") +
    ylab("Cumulative Coverage") +
    ggtitle("Outlier Detection Assessment") +
    scale_colour_viridis_c() +
    scale_size_continuous(range = c(3, 10)) +
    ggrepel::geom_label_repel(
        data = coverage_df,
        aes(label = added_module),
        colour = "black",
        size = 6
    ) 
```

We note the presence of high overlaps between some redundant and non-redundant
modules. This suggests that the redundancy should not be interpreted as a strong
discarding of the module, but as a general recommendation.
```{r outlier_filtering}
outlier_output <- outlier_assessment$outlier_output
modules_stats_summary$is_outlier <- outlier_output[rownames(modules_stats_summary)]
selected_modules <- modules_stats_summary %>%
    filter(is_outlier == "no") %>%
    arrange(median_pseudotime) %>%
    pull(module)
print(modules_stats_summary %>%
    arrange(median_pseudotime, iqr_pseudotime, median_umap_distance) %>%
    select(module, is_outlier))
```

Once the uninformative modules were discarded, Starlng determines the transition
relationship between the remaining gene clusters. This is done using the trajectory
graph: each module is associated with a node on the graph based on the median
position. We then draw edges between nodes that are neighbours on the same
trajectory (based on the removal of the other vertices). If cycles are formed,
we remove the edges between the nodes that are that farthest with respect to
the UMAP space.
```{r module_transition_graph, fig.height = 5}
closest_node_per_module <- get_closest_node_to_module(
    trajectory_object = trajectory_object,
    cell_umap = umap_df[, c("UMAP_1", "UMAP_2")],
    module_expr = module_avg_expression[selected_modules],
    expression_threshold = scale_threshold
)
module_medians <- get_module_centroid(split(module_mask, row(module_mask)), umap_df[, 1:2])
module_adjacency <- get_module_transitions(
    trajectory_object,
    closest_node_per_module,
    start_node = NULL,
    similarity_values = module_medians[selected_modules, , drop = FALSE]
)
plot_module_transitions(
    module_adj_matrix = module_adjacency,
    closest_module = closest_node_per_module
)
```

## Other Module Visualisation Plots

Starlng offers multiple ways to visualise the expression patterns of the
remaining modules. One of them is a cell-resolution heatmap, ordered by the
pseudotime values. This visualisation confirms that most of the data is well
represented by the chosen modules.
```{r module_cell_heatmap, fig.width = 8, fig.height = 8}

generate_cell_heatmap(
  expression_matrix = expr_matrix,
  gene_family_list = module_list[selected_modules],
  metadata_df = data.frame(
    seurat_annotations = mtd_value,
    pseudotime = umap_df$pseudotime,
    row.names = rownames(umap_df)
  ),
  metadata_name = "seurat_annotations",
  apply_scale = TRUE,
  k_smooth = 30,
  cap = 2,
  continuous_colors = grDevices::colorRampPalette(c("blue", "white", "red"))(100)
)
```

Another visualisation relies on fitting regression splines (using GAM) that
describe the evolution of the module average expression over the pseudotime.
This plot also suggests that the second module plays an important role in
defining the transitioning population.
```{r module_pseudotime_curves, fig.width = 7, fig.height = 5}
plot_module_trends_over_pseudotime(
  expression_list = module_avg_expression[selected_modules],
  pseudotime = umap_df$pseudotime,
  module_colours = module_colours
) + theme(legend.position = "none")
```

## Hub Genes

The next step that Starlng follows is to define the impact of each gene in the
overall pattern of expression of the module it belongs to. We defined this as the
product between the fraction to the maximum weight of intra-module edges and a
f1 score that takes into account the overlap between the module-specific population
and the gene-specific population.
For each module, we sort the genes decreasing based on this compound score. The
genes with the highest values act as hubs inside the module.
```{r hub_genes}
inside_module_weights <- get_per_module_weight(
    gene_adj_matrix = gene_adj_matrix,
    gene_modules = module_list
)
hub_score_df <- get_gene_overlap_stat(
  expr_matrix = expr_matrix,
  gene_modules = module_list,
  module_expr = module_avg_expression,
  module_expression_threshold = scale_threshold,
  gene_expression_percentile = 0.5,
  gene_expression_threshold = 0,
  scale = TRUE,
  total_weight = inside_module_weights
)
hub_score_df$module <- chosen_clustering[rownames(hub_score_df)]
hub_score_df$gene <- rownames(hub_score_df)

hub_genes <- hub_score_df %>%
    group_by(module) %>%
    slice_max(combined_score, n = 3) %>%
    filter(module %in% selected_modules)
print(hub_genes)
```

We showcase the position of the top genes on the gene UMAP. In the following plot,
only the top 10% of the edges (ordered by their weights) are shown. Edges between
modules that are not neighbours are discarded.
```{r gene_umap_hubs, fig.width = 8, fig.height = 6}
filtered_gene_adj <- get_filtered_gene_adjacency(
  gene_modules = module_list[selected_modules],
  module_adjacency = module_adjacency,
  gene_adjacency = gene_adj_matrix,
  hub_genes = hub_genes,
  percentage_non_hub_nodes = 1,
  percentage_edges = 0.1
)

plot_gene_hub_umap(
  umap_df = gene_umap,
  filtered_gene_adj = filtered_gene_adj,
  module_colours = module_colours,
  legend_text_size = 10
)
```

## Functional Analysis

Finally, Starlng uses the modules and the gene ordering to profile the functional
interpretation of their associated populations.
The first tool for this type of functional analysis is to perform Enrichment
Analysis. In our context, we use `gprofiler2` to determine the top 3 terms from
the GO:BP database using only the top 15 hub genes from each module.

We remark the presence of terms such as _regulation of natural killer_ for module
11, _T cell receptor signaling_ for module 2 that match the original annotation.
```{r enrichment_analysis, fig.height = 8, fig.width = 8}
hub_genes <- hub_score_df %>%
    group_by(module) %>%
    slice_max(combined_score, n = 15) %>%
    filter(module %in% selected_modules)
go_bp_enrichment <- lapply(selected_modules, function(module_name) {
  used_genes <- hub_genes %>% filter(module == module_name) %>% pull(gene)
  gprofiler2::gost(
    query = used_genes,
    sources = "GO:BP",
    organism = "hsapiens",
    domain_scope = "custom_annotated",
    custom_bg = rownames(expr_matrix),
    correction_method = "fdr"
  )
})
names(go_bp_enrichment) <- selected_modules

plot_enrichment_top_terms(
  enrichment_result = filter_enrichment_results(go_bp_enrichment),
  top_n = 3,
  colour_column = "intersection_size"
)
```

Another approach is to derive the main Transcription Factors associated with
each module. For this, we use `gprofiler2` to compare the gene sets against
the `Transfac` databset. Starlng groups the TFs by their name (attempting to converge
equivalent names such as HNF4a and HNF4-alpha) and shows, in a dotplot, the
association between TFs and modules, along with information about the intersection
size.

We remark the presence of C\\EBP-beta, a Transcription Factor that is associated
to the differention of myleoif and lymphoid cells.
```{r identify_tfs, fig.width = 8, fig.height = 6}
tfs <- get_transcription_factors(
  gene_list = module_list,
  organism = "hsapiens",
  bg_genes = rownames(expr_matrix)
)
tf_stats <- get_tf_stats(tfs, include_intersection_set = TRUE)
tf_stats <- add_tf_hub_stats(tf_stats, hub_genes)

plot_module_tfs_bubbleplot(
  tf_stats = tf_stats
)
```

An alternative is to visualise the interaction between the Transcription Factors
as well as the major regulons in a graph plot.
```{r tf_module_transitions, fig.width=8, fig.height = 6}
tf_module_adj <- get_module_transitions(
  trajectory_object = trajectory_object,
  closest_module = closest_node_per_module[names(tfs)],
  similarity_values = module_medians[names(tfs), , drop = FALSE]
)

tf_gene_graph <- get_tf_gene_network(
  tf_stats = tf_stats,
  module_adjacency = tf_module_adj,
  top_n_factors = 7,
  hub_genes = hub_genes
)

plot_module_tfs_ggraph(
  module_tf_g = tf_gene_graph,
  exclude_non_hub_genes = TRUE
)
```

```{r sessionInfo}
sessionInfo()
```



